---
title: "Code Handout - Data Visualisation with ggplot2"
---

This document contains all of the functions that we have covered thus far in the
course. It will be updated every week, after we've added new skills. Each 
function is presented alongside an example of how it is used. 

All of the examples below are in the context of the Palmer Penguins, found 
[here (link)](https://allisonhorst.github.io/palmerpenguins/index.html). 

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 3, 
                  fig.height = 3, 
                  message = FALSE, 
                  warning = FALSE,
                  eval = FALSE)

```


## Foundations of `ggplot()`

* `ggplot()` -- a function to create the shell of a visualization, where 
specific variables are mapped to different aspects of the plot 

```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species))
```

- `aes()` -- aesthetics that can be used when creating a `ggplot()`, where the 
aesthetics can either be hard coded (e.g. `color = "blue"`) or associated with 
a variable (e.g. `color = sex`). 
  * The following are the aesthetic options for *most* plots:
    + `x`
    + `y`
    + `alpha` -- changes transparency
    + `color` -- produces colored outline 
    + `fill` -- fills with color
    + `group` -- used with categorical variables, similar to color


- **`+`** -- an important aspect creating a `ggplot()` is to note that the 
    `geom_XXX()` function is separated from the `ggplot()` function with a plus
    sign, `+`.
  * `ggplot()` plots are constructed in series of layers, where the plus sign
    separates these layers.  
  * Generally, the `+` sign can be thought of as the end of a line, so you
    should always hit enter/return after it. While it is not mandatory to move
    to the next line for each layer, doing so makes the code a lot easier to
    organize and read. 

```{r, fig.width = 6}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point()
```


## Geometric Objects to Visualize the Data

- `geom_histogram( )` -- adds a histogram to the plot, 
where the observations are binned into ranges of values and then frequencies
of observations are plotted on the y-axis
  * You can specify the number of bins you want with the `bins` argument 
  
```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram(bins = 20)
```

- `geom_boxplot( )` -- adds a boxplot to the plot, where observations are 
aggregated (summarized), the min, Q1, median, Q3, and maximum are plotted as the
box and whiskers, and "outliers" are plotted as points. 
  * You can plot a vertical boxplot by specifying the `x` variable, or a 
  horizontal boxplot by specifying the `y` variable.
  * Note: the min and max may not be included in the whiskers, if they are
  deemed to be "outliers" based on the $1.5 \times \text{IQR}$ rule. 

```{r}
## Horizontal boxplot
penguins %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_boxplot()

## Vertical boxplot
penguins %>% 
  ggplot(aes(y = bill_length_mm)) + 
  geom_boxplot()
```

- `geom_density()` -- adds a density curve to the plot, where the probability
density is plotted on the y-axis (so the density curve has a total area of one). 
  * By default this creates a density curve without shading. By specifying a 
    color in the `fill` argument, the density curve is shaded. 
  * Can be thought of as the "one group" violin plot!
  
```{r, warning=FALSE, message=FALSE}
penguins %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_density(fill = "tomato")
```

- `geom_violin()` -- plots violins for each level of a categorical variable
  * Can be thought of as a hybrid mix of `geom_boxplot()` and `geom_density()`, 
  as the density is displayed, but it is reflected to provide a plot similar in 
  nature to a boxplot. 
  * To obtain violins stacked vertically, declare the categorical variable as `y`.
  To obtain side-by-side violins, declare the categorical variable as `x`.

```{r}
## Stacked vertically
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = species)) + 
  geom_violin()

## Side-by-side
penguins %>% 
  ggplot(aes(y = bill_length_mm, x = species)) + 
  geom_violin()
```

- `geom_bar()` -- creates a barchart of a categorical variable 
  * Can produce stacked barcharts by specifying a variable as the `fill` 
  aesthetic. 
  * Can change from stacked barchart to a side-by-side barchart by specifying 
  `position = "dodge"`. 
  * If your data are already in counts (e.g. output from `count()`), then you 
  can specify the `stat = "identity"` argument inside `geom_bar()`. 

```{r}
## Stacked barchart
penguins %>%
    ggplot(aes(x = species)) +
    geom_bar(aes(fill = sex))

## Side-by-side barchart
penguins %>%
    ggplot(aes(x = species)) +
    geom_bar(aes(fill = sex),
             position = "dodge")

## If data are raw counts
penguins %>% 
  count(species, sex) %>% 
  ggplot(aes(x = species, y = n)) + 
  geom_bar(aes(fill = sex),
           stat = "identity",
           position = "dodge")
```

- `geom_point()` -- plots each observation as an (x, y) point, used to create
scatterplots
  * Can use `alpha` to increase the transparency of the points, to reduce 
  overplotting. 
  * Can specify `aes`thetics inside of `geom_point()` for local aesthetics (point
  level) or inside of `ggplot()` for global aesthetics (plot level)

```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) + 
  geom_point(aes(color = species))
```

- `geom_jitter()` -- plots each observation as an (x, y) point and adds a small 
amount of jitter around the point
  * Useful so that we can see each point in the locations where there are
  overlapping points. 
  * Can specify the `width` and `height` of the jittering using the optional 
  arguments. 
  
```{r}
penguins %>% 
  ggplot(aes(y = body_mass_g, x = species)) + 
  geom_violin() + 
  geom_jitter(aes(color = sex), width = 0.25, height = 0.25)
```

- `geom_smooth()` -- plots a line over a set of points, draws the readers eye 
to a specific trend
  * The methods we will use are "lm" for a linear model (straight line), and 
  "loess" for a wiggly line
  * By default, the smoother gives you gray SE bars, to remove these add 
  `se = FALSE` 
  
```{r, fig.width = 6}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") 
```

- `facet_wrap()` -- creates subplots of your original plot, based on the levels
of the variable you input
  * To facet by one variable, use `~variable`. 
  * To facet by two variables, use `variable1 ~ variable2`. 
  * If you prefer for your facets to be organized in rows or columns, use the 
  `nrow` and/or `ncol` arguments. 
  
```{r, fig.width = 12}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~island, nrow = 1)
```


## Plot Characteristics 

- `labs()` -- specifies the plot labels, possible labels are: x, y, color, fill, 
title, and subtitle

```{r, fig.width = 6}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "Bill Length (mm)", 
       y = "Bill Depth (mm)", 
       color = "Penguin Species")
```

- `theme_bw()` -- changes the plotting background to the classic dark-on-light
ggplot2 theme.
  * This theme may work better for presentations displayed with a projector.
  * Other theme options are `theme_minimal()`, `theme_light()`, and `theme_void()`. 
  
```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "Bill Length (mm)", 
       y = "Bill Depth (mm)", 
       color = "Penguin Species") + 
  theme_bw()
```

- `theme()` -- 
  * Possible options are:
    + `panel.grid` -- controls the grid lines (`panel.grid = element_blank()` 
    removes grid lines)
    + `text` -- specifies font size for the entire plot (e.g. 
    `text = element_text(size = 16)` 
    + `axis.text.x` -- specifies the font size for the x-axis text
    + `axis.text.y` -- specifies the font size for the y-axis text
    + `plot.title` -- specifies aspects of the plot title, can use 
    `plot.title = element_text(hjust = 0.5)` to centre the title
    
```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "Bill Length (mm)", 
       y = "Bill Depth (mm)", 
       color = "Penguin Species") + 
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12))
```


## Exporting Plots 

- `ggsave()` -- convenient function for saving a plot 
  * Unless specified, defaults to the last plot that was made. 
  * Uses the size of the current graphics device to determine the size of the 
  plot. 

```{r}
plot1 <- penguins %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~island, nrow = 1)

ggsave(path = "images/faceted_plot.png", plot = plot1)
```

