---
title: "Code Handout"
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 3, 
                  fig.height = 3, 
                  message = FALSE, 
                  warning = FALSE,
                  eval = FALSE)

```

This document contains all of the functions that we have covered thus far in the
course. It will be updated every week, after we've added new skills. Each 
function is presented alongside an example of how it is used. 

All of the examples below are in the context of the Palmer Penguins, found 
[here (link)](https://allisonhorst.github.io/palmerpenguins/index.html). 



## Packages

- `library()` -- loads packages into your `R` session
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```



## Inspecting Data

- `glimpse()` -- shows a summary of the dataset, the number of rows and columns, 
variable names, and the first 10 entries of each variable 
```{r}
glimpse(penguins)
```



## Working with Data

- `<-` -- "assignment arrow", assigns a value (vector, dataframe, single value)
to the name of a variable 
```{r}
penguins_2007 <- penguins %>% 
  filter(year == 2007)
```



- `c()` -- the "concatenate" function combines inputs to form a vector, the 
values have to be the same data type. 

```{r}
cat_variables <- c("Species", "Island", "Sex")
```

\newpage

## Verbs of Data Wrangling

- `select()` -- selects variables (columns) from a dataframe 
```{r}
penguins %>% 
select(species)
```

- `filter()` -- filters observations (rows) out of / into a dataframe, where
the inputs (arguments) are the conditions to be satisfied in the data that are
kept 
```{r}
## It's nice to have a new line for each condition, so your code is easier to read!
penguins %>% 
filter(species == "Adelie",
       body_mass_g > 3000,
       year == 2008)

```

**Logical operators:** Filtering for certain observations (e.g. flights from a
particular airport) is often of interest in data frames where we might want to
examine observations with certain characteristics separately from the rest of
the data. To do so, you can use the `filter` function and a series of **logical
operators**. The most commonly used logical operators for data analysis are as
follows:

-   `==` means "equal to"
-   `!=` means "not equal to"
-   `>` or `<` means "greater than" or "less than"
-   `>=` or `<=` means "greater than or equal to" or "less than or equal to"

- `mutate()` -- creates new variables or modifies existing variables
```{r}
penguins %>% 
  filter(is.na(bill_length_mm) != TRUE, 
         is.na(bill_depth_mm) != TRUE) %>% 
  mutate(body_mass_kg = body_mass_g / 1000)
```


- `group_by()` -- groups the dataframe based on levels of a categorical variable, 
usually used alongside `summarize()`

```{r, eval = FALSE}
penguins %>% 
  group_by(island)
```


- summarize()` -- creates data summaries of variables in a dataframe, for grouped 
summaries use alongside `group_by()`

```{r}
penguins %>% 
  filter(is.na(body_mass_g) != TRUE) %>% 
  group_by(island) %>% 
  summarize(mean_mass = mean(body_mass_g))  

```


- `ungroup()` -- removes the grouping of a dataframe, typically used after group
summaries when additional ungrouped operations are required 
```{r}
penguins %>% 
  filter(is.na(body_mass_g) != TRUE) %>% 
  group_by(island) %>% 
  summarize(mean_mass = mean(body_mass_g)) %>% 
  ungroup() 
```


- `arrange()` -- orders a dataframe based on the values of a numerical variable, 
paired with `desc()` to order in descending order

```{r}
penguins %>% 
  filter(is.na(body_mass_g) != TRUE) %>% 
  group_by(island) %>% 
  summarize(mean_mass = mean(body_mass_g)) %>% 
  arrange(desc(mean_mass))
```


- `%>%` -- the "pipe" operator, joins sequences of data wrangling steps together, 
works with any function that has `data = ` as the first argument 
```{r}
penguins %>%
  select(species, island, body_mass_g, sex, year) %>% 
  filter(island ==   "Torgersen", 
         is.na(body_mass_g) != TRUE) %>% 
  group_by(species, year) %>% 
  summarize(mean_mass = mean(body_mass_g),
            median_mass = median(body_mass_g),
            observations = n()) %>% 
  arrange(desc(mean_mass))
```


## Other Data Wrangling Tools
  
- `count()` -- counts the number of observations (rows) of the different levels 
of a categorical variable
  * can add `sort = TRUE` to sort the table in descending order (similar to 
  using `arrange(desc())` )
```{r}
penguins %>% 
count(species)
```

- `mean()` -- finds the mean of a numerical variable, not resistant to `NA` values, 
so either filter out prior or use `na.omit = TRUE` argument
  * Other summary functions include:
    +  `var()` -- find the variance of a numerical variable
    +  `sd()` -- finds the standard deviation of a numerical variable
    + `IQR()` -- find the innerquartile range (Q3 - Q1) of a numerical variable
    + `median()` -- finds the median of a numerical variable 

- `is.na()` -- returns a vector of `TRUE` and `FALSE` values corresponding to 
whether a particular row of a variable was `NA` (missing)
```{r}
penguins %>% 
  mutate(missing_weight = is.na(body_mass_g))
```

- `sample_n()` -- selects $n$ rows from the dataframe, based on the value of 
`size` specified
```{r}
penguins %>% 
  sample_n(size = 10)
```

- `replace_na()` -- replaces NA values with the value specified 
  * The values to be replaced must be passed to the function (input) as a 
  `list()` object. 
```{r}
penguins %>% 
  replace_na(list(bill_length_mm = "no_measurement", 
                  bill_depth_mm = "no_measurement")) %>% 
  glimpse()
```

- `separate_rows()` -- separates a variable with multiple values based on the 
delimiter specified. 
  * Variables whose entries are stored as a list with commas or semicolons are 
  great candidates for this function!
  
- `rowSums()` -- forms row sums for numeric variables 
  * Note: In the lesson `rowSums()` was used on a `logical` variable, because
  logical values can be numerically represented as 0 (FALSE) and 1 (TRUE)
```{r}
x <- tibble(x1 = 3, x2 = c(4:1, 2:5))
rowSums(x)
```


## Pivoting Dataframes

- `pivot_wider()` -- transforms a dataframe from long to wide format
  * takes three principal arguments: 
    1. the data
    2. the *names_from* column variable whose values will become new column names
    3. the *values_from* column variable whose values will fill the new column
   variables.
  * Further arguments include `values_fill` which, if set, fills in missing
  values with the value provided.
  
```{r}
wide <- penguins %>% 
  mutate(island_logical = TRUE) %>% 
  pivot_wider(names_from = species, 
              values_from = island_logical, 
              values_fill = list(island_logical = FALSE))

glimpse(wide)
```

- `pivot_longer()` -- transforms a dataframe from wide to long format
  * takes four principal arguments:
  1. the data
  2. *cols* are the names of the columns we use to fill the a new values variable
   (or to drop).
  3. the *names_to* column variable we wish to create from the *cols* provided.
  4. the *values_to* column variable we wish to create and fill with values
  associated with the *cols* provided.
  
```{r}
wide %>% 
  pivot_longer(cols = Adelie:Gentoo, 
               names_to = "species", 
               values_to = "island_logical")
```

## Extracting Data

- `write_csv()` -- writes a dataframe to a csv file, output into the file path 
specified

```{r}
write_csv(wide, path = "data/penguins_wide.csv")
```

