---
title: "Code Handout - Introduction to R"
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 3, 
                  fig.height = 3, 
                  message = FALSE, 
                  warning = FALSE,
                  eval = FALSE)

```

This document contains all of the functions that were covered in the 
_Introduction to R_ workshop. Each function is presented alongside an example of
how it can be used. 


## Creating Objects 

- `<-` -- "assignment arrow", assigns a value (vector, dataframe, single value)
to the name of a variable 
```{r}

x <- 3
y <- c(1, 2, 3)
z <- x + y
```

- `c()` -- the "concatenate" function combines inputs to form a vector, the 
values have to be the same data type. 
```{r}
animals <- c("bird", "cat", "dog")
numbers <- c(1, 14, 57, 89)
logicals <- c(TRUE, FALSE, TRUE, TRUE)
```


## Inspecting Objects

- `str()` -- compact display of the structure of an R object 
```{r}
str(animals)
```

- `class()` -- returns the type of element of any R object
```{r}
class(logicals)
```

- `typeof()` -- returns the data type or storage mode of any R object
```{r}
typeof(numbers)
```


## Functions in R 

- `args()` -- returns the arguments of a function
```{r}
args(round)
```

- named arguments -- the name of the argument the function expects
  * You can choose to not name your arguments, __if__ you know the __exact__ 
  order they should be in! 
  * However, we generally discourage this.
```{r}
## Either of these work, since the digits argument is named explicitly.
round(3.14159, digits = 2)
round(digits = 2, 3.14159)

## This does not work, since the arguments are not named and in the incorrect order. 
round(2, 3.14159)
```


## Functions to Summarize Data

- `sqrt()` -- returns the square root of a numeric variable
```{r}
sqrt(numbers)
```

- `mean()` -- returns the mean of a numeric variable
  * You can add the `na.rm` argument, to remove `NA` values before calculating 
  the mean.
```{r}
sqrt(numbers)
```

- `max()` -- returns the maximum of a numeric variable
  * You can add the `na.rm` argument, to remove `NA` values before calculating 
  the max.
```{r}
sqrt(numbers)
```

- `sum()` -- returns the sum of a numeric variable
  * You can add the `na.rm` argument, to remove `NA` values before calculating 
  the sum.
```{r}
sqrt(numbers)
```

- `length()` -- returns the length of a vector (of any datatype)
```{r}
length(animals)
```


## Subsetting Data

- `[]` -- used to subset elements from a vector
```{r}
animals[3]
## selects the third element

animals[2:3]
## selects the second and third element

animals[c(1, 3)]
## selects the first and third element
```

- relational operators -- return logical values indicating where a relation is 
satisfied. The most commonly used logical operators for data analysis are as follows:
  * `==` means "equal to"
  * `!=` means "not equal to"
  * `>` or `<` means "greater than" or "less than"
  * `>=` or `<=` means "greater than or equal to" or "less than or equal to"
```{r}
animals == "dog"

animals != "cat"

numbers > 4

numbers <= 12
```


- logical operators -- join subset criteria together 
  * `&` means "and" -- where two criteria must **both** be satisfied
  * `|` means "or" -- where at least one criteria must be satisfied
```{r}
numbers > 4 & numbers < 20

animals == "dog" | animals == "cat"
```
  

- `%in%` -- the "inclusion operator", allows you to test if any of the elements
of a search vector (on the left hand side) are found in the target vector (on
the right hand side). 
  * The levels of the target vector must be included in a vector (`c()`). 
```{r}
possessions <- c("car", "bicycle", "radio", "television", "mobile_phone")

possessions %in% c("car", "bicycle", "motorcycle")
```


## Missing Data 

- `is.na()` -- returns a vector of logical values indicating which elements of 
a vector have `NA` values
  * Often combined with `!`, where the `!` negates the previous statement (e.g. 
  `!TRUE` is equal to `FALSE`). 
```{r}
missing <- c(1, 3, NA, 7, 12, NA)

is.na(missing)

!is.na(missing)
```

- `na.omit()` -- removes the observations with `NA` values
```{r}
na.omit(missing)
```

- `complete.cases()` -- returns a vector of logical values indicating which
elements of a vector **are not** missing (`NA`) values
```{r}
complete.cases(missing)
```

