---
title: "Code Handout - Starting with Data"
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 3, 
                  fig.height = 3, 
                  message = FALSE, 
                  warning = FALSE,
                  eval = FALSE)

```

This document contains all of the functions that were covered in the 
_Introduction to R_ workshop. Each function is presented alongside an example of
how it can be used. 

All of the examples below are in the context of the Palmer Penguins, found 
[here (link)](https://allisonhorst.github.io/palmerpenguins/index.html). 


## Packages

- `library()` -- loads packages into your `R` session
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```


## Importing Data
- `read_csv()` -- function to import a csv file. 
  * First argument is the path to the data, passed as a character 
  (inside quotations). 
  * You can specify what values should be considered missing, using the `na`
  argument. 
```{r}
penguins <- read_csv("data/penguins.csv")
```


## Inspecting Data

- `dim()` - returns a vector with the number of rows as the first element,
and the number of columns as the second element (the **dim**ensions of
the object)
```{r}
dim(penguins)
```

- `nrow()` - returns the number of rows
- `ncol()` - returns the number of columns
```{r}
nrow(penguins)
ncol(penguins)
```

- `head()` - displays the first 6 rows of the dataframe
- `tail()` - displays the last 6 rows of the dataframe
```{r}
head(penguins)
tail(penguins)
```

- `names()` - returns the all of the names of an object (both row and column) 
- `colnames()` - returns column names for dataframes (without row names)
```{r}
names(penguins)
colnames(penguins)
```

- `glimpse()` - provides a preview of the data, where column names are presented 
with their associated data types, and the entries from each column are printed 
in each row 
```{r}
glimpse(penguins)
```

- `str()` - returns the structure of the object and information about the class,
the names and data types of each column, and a preview of the first entries of
each column
```{r}
str(penguins)
```

- `summary()` - provides summary statistics for each column
  * Note: summary statistics for character variables are not meaningful, as they 
  simply state the number of observations (length) of the variable
```{r}
summary(penguins)
```


## Subsetting Data

- `[]` -- selects rows and columns from a dataframe
  * The first entry is the row number, the second entry is the column number(s),
  and they are separated with a comma. 
```{r}
## Selects the element in the first row, second column
penguins[1, 2]

## Selects every element in the fourth row
penguins[4, ]

## Selects every element in the third column
penguins[, 3]
```

- `[[]]` -- selects a column from a dataframe
  * Inside the brackets you can pass either the number of the column or the 
  name of the column (in quotations)
```{r}
penguins[[1]]

penguins[["island"]]
```

- `$` -- selects a column from a dataframe, where the name of the dataframe is 
on the left and the name of the column is on the right
```{r}
penguins$body_mass_g
```


## Working with Different Data Types

- `factor()` -- creates a categorical variable from a character or numeric 
variable, variable has a factor datatype
  * the values (level) of the factor levels is specified in the `levels` 
  argument, where the levels must be specified in a vector (using `c()`)
  * Note: the order you wish for the levels to appear is how you should list 
  them in the `levels` argument, you can also specify `ordered = TRUE` to 
  ensure the levels remain in this order
```{r}
penguins$year_fct <- factor(penguins$year, 
                               levels = c("2007", "2008", "2009"), 
                            ordered = TRUE)
```

- `as.factor()` -- creates a categorical variable from a character or numeric 
variable, variable has a factor datatype
  * does not allow for you to specify the order of the levels
  * defaults to alphabetical ordering for factor levels
```{r}
penguins$year_fct <- as.factor(penguins$year)
```

- `levels()` -- returns the levels of a variable with a factor datatype, in the 
order they were stored
  * Note: this function will not work for character datatypes
```{r}
levels(penguins$year_fct)
```

- `nlevels()` -- returns the number of levels of a variable with a factor
datatype
  * Note: this function will not work for character datatypes
```{r}
nlevels(penguins$year_fct)
```

- `as.character()` -- creates a character variable from a numeric or factor 
variable
```{r}
penguins$species_chr <- as.character(penguins$species)
```

- `ymd()` -- transforms dates stored as character or numeric variables to dates
  * Note: to use this function, dates must be stored in year-month-day format
  * The function does well with heterogeneous formats (as seen below), but
  formats where some of the entries are not in double digits may not be parsed 
  correctly.
```{r}
x <- c("2009-01-01", "2009-01-02", "2009-01-03")
ymd(x)
```

- `day()` -- extracts the day (number) of a date variable 
```{r}
day(x)
```

- `month()` -- extracts the month (number) of a date variable
```{r}
month(x)
```

- `year()` -- extracts the year of a date variable
```{r}
year(x)
```


## Visualizing Data

- `plot()` -- a generic function for plotting R objects
  * In this lesson `plot()` was used to create bargraphs of categorical
  variables.
```{r}
plot(penguins$species)
```


