---
title: "Starting with Data"
teaching: 50
exercises: 30
questions:
- "What is a data.frame?"
- "How can I read a complete csv file into R?"
- "How can I get basic summary information about my dataset?"
- "How can I change the way R treats strings in my dataset?"
- "Why would I want strings to be treated differently?"
- "How are dates represented in R and how can I change the format?"
objectives:
- "Describe what a data frame is."
- "Load external data from a .csv file into a data frame."
- "Summarize the contents of a data frame."
- "Describe the difference between a factor and a string."
- "Convert between strings and factors."
- "Reorder and rename factors."
- "Change how character strings are handled in a data frame."
- "Examine and change date formats."

keypoints:
- "Use read.csv to read tabular data in R."
- "Use factors to represent categorical data in R."
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
source("../bin/download_data.R")
```

## Presentation of the SAFI Data

SAFI (Studying African Farmer-Led Irrigation) is a study looking at farming
and irrigation methods in Tanzania and Mozambique. The survey data
was collected through interviews conducted between November 2016 and
June 2017. For this lesson, we will be using a subset of the
available data. For information about the full teaching dataset used
in other lessons in this workshop, see the
[dataset description](http://www.datacarpentry.org/socialsci-workshop/data/).

We will be using a subset of the cleaned version of the dataset that
was produced through cleaning in OpenRefine. Each row holds
information for a single interview respondent, and the columns
represent:

| column_name | description |
| ----------- | ----------- |
| key_id | Added to provide a unique Id for each observation. (The InstanceID field does this as well but it is not as convenient to use) |
| village | Village name |
| interview_date | Date of interview |
| no_membrs | How many members in the household? |
| years_liv | How many years have you been living in this village or neighboring village? |
| respondent_wall_type | What type of walls does their house have (from list) |
| rooms | How many rooms in the main house are used for sleeping? |
| memb_assoc | Are you a member of an irrigation association? |
| affect_conflicts | Have you been affected by conflicts with other irrigators in the area? |
| liv_count | Number of livestock owned. |
| items_owned | Which of the following items are owned by the household? (list) |
| no_meals | How many meals do people in your household normally eat in a day? |
| months_lack_food | Indicate which months, In the last 12 months have you faced a situation when you did not have enough food to feed the household? |
| instanceID | Unique identifier for the form data submission |


You are going load the data in R's memory using the function `read_csv()` 
from the **`readr`** package, which is part of the **`tidyverse`**; learn 
more about the **`tidyverse`** collection of packages 
[here](https://www.tidyverse.org/).
**`readr`** gets installed as part as the **`tidyverse`** installation. 
When you load  the **`tidyverse`** (`library(tidyverse)`), the core packages 
(the packages used in most data analyses) get loaded, including **`readr`**. 

So, before we can use the `read_csv()` function, we need to load the package. 
Also, if you recall, the missing data is encoded as "NULL" in the dataset. 
We'll tell it to the function, so R will automatically convert all the "NULL" 
entries in the dataset into `NA`.

```{r, eval=TRUE,  message=FALSE, purl=FALSE}
library(tidyverse)
interviews <- read_csv("data/SAFI_clean.csv", na = "NULL")
```

This statement creates a data frame but doesn't show any data because, as you might recall,
assignments don't display anything. (Note, however, that `read_csv` may show informational
text about the data frame that is created.) If we want to check that our data has been
loaded, we can see the contents of the data frame by typing its name: `interviews`.

```{r, results='hold', purl=FALSE}
interviews
## Try also
## View(interviews)
## head(interviews)
```

> ## Note
>
> `read_csv()` assumes that fields are delimited by commas. However, in several
> countries, the comma is used as a decimal separator and the semicolon (;) is
> used as a field delimiter. If you want to read in this type of files in R,
> you can use the `read_csv2` function. It behaves exactly like `read_csv` but
> uses different parameters for the decimal and the field separators. If you are
> working with another format, they can be both specified by the user. Check out
> the help for `read_csv()` by typing `?read_csv` to learn more. There is also
> the `read_tsv()` for tab-separated data files, and `read_delim()` allows you
> to specify more details about the structure of your file.
{: .callout}

## What are data frames and tibbles?

Data frames are the _de facto_ data structure for tabular data in `R`, and what 
we use for data processing, statistics, and plotting.

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Data frames are analogous to
the more familiar spreadsheet in programs such as Excel, with one key difference.
Because columns are vectors,
each column must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](../fig/data-frame.svg)

A data frame can be created by hand, but most commonly they are generated by the
functions `read_csv()` or `read_table()`; in other words, when importing
spreadsheets from your hard drive (or the web).

A tibble is an extension of `R` data frames used by the **`tidyverse`**. When the 
data is read using `read_csv()`, it is stored in an object of class `tbl_df`, 
`tbl`,  and `data.frame`. You can see the class of an object with

```{r results='hold', purl=FALSE}
class(interviews)
```

As a `tibble`, the type of data included in each column is listed in an 
abbreviated fashion below the column names. For instance, here `key_ID` is a 
column of integers (abbreviated `<int>`), `village` is a column of characters 
(`<chr>`) and the `interview_date` is a column in the "date and time" format 
(`<dttm>`).

## Inspecting data frames

When calling a `tbl_df` object (like `interviews` here), there is already a lot of information about our data frame being displayed such as the number of rows, the number of columns, the names of the columns, and as we just saw the class of data stored in each column. However, there are functions to extract this information from data frames.  Here is a non-exhaustive list of some of these
functions. Let's try them out!

Size:
* `dim(interviews)` - returns a vector with the number of rows as the first element,
and the number of columns as the second element (the **dim**ensions of
the object)
* `nrow(interviews)` - returns the number of rows
* `ncol(interviews)` - returns the number of columns

Content:
* `head(interviews)` - shows the first 6 rows
* `tail(interviews)` - shows the last 6 rows

Names:
* `names(interviews)` - returns the column names (synonym of `colnames()` for
`data.frame` objects)

Summary:
* `str(interviews)` - structure of the object and information about the class,
length and content of each column
* `summary(interviews)` - summary statistics for each column

Note: most of these functions are "generic." They can be used on other types of
objects besides data frames.



## Indexing and subsetting data frames

Our `interviews` data frame has rows and columns (it has 2 dimensions). If we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to results with
different classes.


```{r, purl=FALSE}
## first element in the first column of the data frame (as a vector)
interviews[1, 1]
## first element in the 6th column (as a vector)
interviews[1, 6]
## first column of the data frame (as a vector)
interviews[[1]]
## first column of the data frame (as a data.frame)
interviews[1]
## first three elements in the 7th column (as a vector)
interviews[1:3, 7]
## the 3rd row of the data frame (as a data.frame)
interviews[3, ]
## equivalent to head_interviews <- head(interviews)
head_interviews <- interviews[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
interviews[, -1]          # The whole data frame, except the first column
interviews[-c(7:131), ]   # Equivalent to head(interviews)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
interviews["village"]       # Result is a data frame
interviews[, "village"]     # Result is a data frame
interviews[["village"]]     # Result is a vector
interviews$village          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

> ## Exercise
>
> 1. Create a data frame (`interviews_100`) containing only the data in
>    row 100 of the `interviews` dataset.
>
> 2. Notice how `nrow()` gave you the number of rows in a data frame?
>
>      * Use that number to pull out just that last row in the data frame.
>      * Compare that with what you see as the last row using `tail()` to make
>        sure it's meeting expectations.
>      * Pull out that last row using `nrow()` instead of the row number.
>      * Create a new data frame (`interviews_last`) from that last row.
>
> 3. Use `nrow()` to extract the row that is in the middle of the data frame.
>    Store the content of this row in an object named `interviews_middle`.
>
> 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
>    `head(interviews)`, keeping just the first through 6th rows of the interviews
>    dataset.
>
> > ## Solution
> >
> > ```{r}
> > ## 1.
> > interviews_100 <- interviews[100, ]
> > ## 2.
> > # Saving `n_rows` to improve readability and reduce duplication
> > n_rows <- nrow(interviews)
> > interviews_last <- interviews[n_rows, ]
> > ## 3.
> > interviews_middle <- interviews[(n_rows / 2), ]
> > ## 4.
> > interviews_head <- interviews[-(7:n_rows), ]
> > ```
> {: .solution}
{: .challenge}


## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered (ordinal variable) or unordered (nominal variable). 
While factors look (and often behave) like character vectors, they are actually 
treated as integer vectors by R.
So you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
respondent_floor_type <- factor(c("earth", "cement", "cement", "earth"))
```

R will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e`, even though the first element in this vector is
`"earth"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(respondent_floor_type)
nlevels(respondent_floor_type)
```

Sometimes, the order of the factors does not matter. Other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high").
It may improve your visualization, or it may be required by a particular type of
analysis. Here, one way to reorder our levels in the `respondent_floor_type` vector would be:

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order
respondent_floor_type <- factor(respondent_floor_type, levels = c("earth", "cement"))
respondent_floor_type # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"cement"`,
`"earth"` is more descriptive than `1`, and `2`. Which one is "earth"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and need to recode "cement" to "brick".

```{r, results=TRUE, purl=FALSE}
levels(respondent_floor_type)
levels(respondent_floor_type)[2] <- "brick"
levels(respondent_floor_type)
respondent_floor_type
```

So far, your factor is unordered, like a nominal variable. R does not know the 
difference between a nominal and an ordinal variable. You make your factor an
ordered factor by using the `ordered=TRUE` option insude your factor function.
Note how the reported levels changed from the unordered factor above to the
ordered version below. Ordered levels use the less than sign `<` to denote
level ranking.

```{r, results=TRUE, purl=FALSE}
respondent_floor_type_ordered = factor(respondent_floor_type, ordered=TRUE)
respondent_floor_type_ordered # after setting as ordered factor
```


### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to numbers.
Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)                     # Wrong! And there is no warning...
as.numeric(as.character(year_fct))       # Works...
as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(year_fct)`
* We convert these levels to numeric values using `as.numeric(levels(year_fct))`
* We then access these numeric values using the underlying integers of the
vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `memb_assoc` column from our data frame, convert it into a
factor, and use it to look at the number of interview respondents who were or
were not members of an irrigation association:

```{r factor-plot-default-order, purl=TRUE}
## create a vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc
## convert it into a factor
memb_assoc <- as.factor(memb_assoc)
## let's see what it looks like
memb_assoc
## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see that in
addition to "no"s and "yes"s, there are some respondents for which the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. They do not appear on the plot.
Let's encode them differently so they can counted and visualized in our plot.


```{r factor-plot-reorder, purl=TRUE}
## Let's recreate the vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc
## replace the missing data with "undetermined"
memb_assoc[is.na(memb_assoc)] <- "undetermined"
## convert it into a factor
memb_assoc <- as.factor(memb_assoc)
## let's see what it looks like
memb_assoc
## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```


> ## Exercise
>
> * Rename the levels of the factor to have the first letter in uppercase:
>   "No","Undetermined", and "Yes". 
>
> * Now that we have renamed the factor level to "Undetermined", can you
>   recreate the barplot such that "Undetermined" is last (after "Yes")?
>
> > ## Solution
> >
> > ```{r factor-plot-exercise}
> > levels(memb_assoc) <- c("No", "Undetermined", "Yes")
> > memb_assoc <- factor(memb_assoc, levels = c("No", "Yes", "Undetermined"))
> > plot(memb_assoc)
> ```
> {: .solution}
{: .challenge}


## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. As a reminder from earlier in this lesson, the best
practice for dealing with date data is to ensure that each component of your
date is stored as a separate variable. In our dataset, we have a
column `interview_date` which contains information about the
year, month, and day that the interview was conducted. Let's
convert those dates into three separate columns.

```{r, eval=FALSE, purl=FALSE}
str(interviews)
```
We are going to use the package **`lubridate`**, which is included in the
**`tidyverse`** installation but not loaded by default, so we have to load 
it explicitly with `library(lubridate)`.

Start by loading the required package:

```{r load-package, message=FALSE, purl=FALSE}
library(lubridate)
```

The lubridate function `ymd()` takes a vector representing year, month, and day, and converts it to a
`Date` vector. `Date` is a class of data recognized by R as being a date and can
be manipulated as such. The argument that the function requires is flexible,
but, as a best practice, is a character vector formatted as "YYYY-MM-DD".


Let's extract our `interview_date` column and inspect the structure:

```{r, purl=FALSE}
dates <- interviews$interview_date
str(dates)
```

When we imported the data in R, `read_csv()` recognized that this column contained date information. We can now use the `day()`, `month()` and  `year()` functions to extract this information from the date, and create new columns in our data frame to store it:


```{r, purl=FALSE}
interviews$day <- day(dates)
interviews$month <- month(dates)
interviews$year <- year(dates)
interviews
```

Notice the three new columns at the end of our data frame.

{% include links.md %}
