---
title: "Introducing dplyr and tidyr"
teaching: 30
exercises: 20
questions:
- "How can I select specific rows and/or columns from a data frame?"
- "How can I combine multiple commands into a single command?"
- "How can create new columns or remove existing columns from a data frame?"
- "How can I reformat a dataframe to meet my needs?"
objectives:
- "Describe the purpose of the **`dplyr`** and **`tidyr`** packages."
- "Select certain columns in a data frame with the **`dplyr`** function `select`."
- "Select certain rows in a data frame according to filtering conditions with the **`dplyr`** function `filter` ."
- "Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`."
- "Add new columns to a data frame that are functions of existing columns with `mutate`."
- "Use the split-apply-combine concept for data analysis."
- "Use `summarize`, `group_by`, and `count` to split a data frame into groups of observations, apply a summary statistics for each group, and then combine the results."
- "Describe the concept of a wide and a long table format and for which purpose those formats are useful."
- "Describe what key-value pairs are."
- "Reshape a data frame from long to wide format and back with the `spread` and `gather` commands from the **`tidyr`** package.
- "Export a data frame to a .csv file."
keypoints:
- "Use the `dplyr` package to manipulate dataframes."
- "Use `select()` to choose variables from a dataframe."
- "Use `filter()` to choose data based on values."
- "Use `group_by()` and `summarize()` to work with subsets of data."
- "Use `mutate()` to create new variables."
- "Use the `tidyr` package to change the layout of dataframes."
- "Use `gather()` to go from wide to long format."
- "Use `spread()` to go from long to wide format."
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
interviews <- read.csv("data/interviews.csv")
```

# Data Manipulation using **`dplyr`** and **`tidyr`**

Bracket subsetting is handy, but it can be cumbersome and difficult to read,
especially for complicated operations. Enter **`dplyr`**. **`dplyr`** is a package for
making tabular data manipulation easier. It pairs nicely with **`tidyr`** which enables you to swiftly convert between different data formats for plotting and analysis.

Packages in R are basically sets of additional functions that let you do more
stuff. The functions we've been using so far, like `str()` or `data.frame()`,
come built into R; packages give you access to more of them. Before you use a
package for the first time you need to install it on your machine, and then you
should import it in every subsequent R session when you need it. You should
already have installed the **`tidyverse`** package. This is an
"umbrella-package" that installs several packages useful for data analysis which
work together well such as **`tidyr`**, **`dplyr`**, **`ggplot2`**, **`tibble`**, etc.


The **`tidyverse`** package tries to address 3 common issues that arise when
doing data analysis with some of functions that come with R:

1. The results from a base R function sometimes depend on the type of data.
2. Using R expressions in a non standard way, which can be confusing for new
learners.
3. Hidden arguments, having default operations that new learners are not aware
of.

We have seen in our previous lesson that when building or importing a data frame, the columns that contain characters (i.e., text) are coerced (=converted) into the `factor` data type. We had to set **`stringsAsFactors`** to **`FALSE`** to avoid this hidden argument to convert our data type.

This time will use the **`tidyverse`** package to read the data and avoid having to set **`stringsAsFactors`** to **`FALSE`**

To load the package type:


```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library("tidyverse")
```

## What are **`dplyr`** and **`tidyr`**?

The package **`dplyr`** provides easy tools for the most common data manipulation
tasks. It is built to work directly with data frames, with many common tasks
optimized by being written in a compiled language (C++). An additional feature is the
ability to work directly with data stored in an external database. The benefits of
doing this are that the data can be managed natively in a relational database,
queries can be conducted on that database, and only the results of the query are
returned.

This addresses a common problem with R in that all operations are conducted
in-memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can connect to a database of many hundreds of GB, conduct queries on it directly, and pull
back into R only what you need for analysis.

The package **`tidyr`** addresses the common problem of wanting to reshape your data for plotting and use by different R functions. Sometimes we want data sets where we have one row per measurement. Sometimes we want a data frame where each measurement type has its own column, and rows are instead more aggregated groups. Moving back and forth between these formats is nontrivial, and **`tidyr`** gives you tools for this and more sophisticated  data manipulation.

To learn more about **`dplyr`** and **`tidyr`** after the workshop, you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) and this [one about **`tidyr`**](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf).

We'll read in our data using the `read_csv()` function, from the tidyverse package **`readr`**, instead of `read.csv()`.


```{r, results = 'hide', purl = FALSE}
interviews <- read_csv("data/interviews.csv")

## inspect the data
str(interviews)

## preview the data
# View(interviews)
```

Notice that the class of the data is now `tbl_df`
